\documentclass[a4paper]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}
\usepackage{bbm}
\usepackage{wrapfig}
\usepackage{tabularx}
\usepackage[autostyle=true]{csquotes}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\title{Lecture 5}
\author{Gidon Rosalki}
\date{2025-11-02}


\begin{document}
\maketitle

Wow, this is a little light on content. This was a very biological lesson, with all the concepts in Hebrew. I spent most
of it a little lost.

\section{Basic concepts}\label{sec:basic_concepts} % (fold)

\subsection{Biological systems}\label{sec:biological_systems} % (fold)
What is a system? A system is a group of joined components, working together to a common goal. A \textit{dynamic} system
is a system that changes with time. An example of a dynamic system is Conway's Game of Life, where each cell is not
particularly interesting in and of itself, but together create a much larger and more interesting system. \\ 

A dynamic \textit{biological} system has various requirements to be considered such. It is a collection of parts that
operate together to enable the requirements for life: \begin{itemize}
    \item To collect information / energy / construction material 
    \item To make choices 
    \item To run plans 
    \item To make adjustments 
    \item To grow / reproduce 
    \item To be protected from the surroundings
\end{itemize}

Models can strengthen, and explain known, and unknown properties. They can also decide unknown hypotheses, and provide
control over properties of the system, providing the ability to change them according to our needs. Finally, it can help
identify missing knowledge, which we can then measure in an experiment. 

% subsection Biological systems (end)
% section Basic concepts (end)

\section{Integrated models of dynamic biological systems}\label{sec:integrated_models_of_dynamic_biological_systems} % (fold)
To model is the process of creating new knowledge, from old knowledge. This can be a continuous process, where this new
knowledge is then fed back into creating a new model. Let us consider a cell filter, size on the order of $0.1\mu m$.
This allows through passive diffusion the entry of small inert macromolecules, ions, and water to diffuse passively
through it into the cell, but blocks the large, inert macromolecules. This is incredibly impressive, such a small
filter, with so much capability. The question is though, how does this work? Since we do not have a camera that can tell
us, we need to create an integrative model. Firstly, we collect information about the system, through things like
experiments, physics, intuition, statistics, and so on. Next we create a computerised mathematical model of the system,
as we understand it right now. Next we find a good model of the system, and finally measure the correctness, and
usefulness of this model. This is then fed back into the first step to try and create a more accurate model. 

% section Integrated models of dynamic biological systems (end)

\end{document}
